---
title: "Web design for Shiny developers"
subtitle: "HTML, CSS, Bootstrap, and `bslib`"
author: "Leslie Emery"
format: 
  revealjs:
    smaller: true
    scrollable: true
    theme: dark
---

## Topics

- HTML
- CSS
- Bootstrap
- How Shiny uses html, CSS, and Bootstrap
- bslib


## HTML (1993)

- HTML = "HyperText Markup Language"
- Markup language = formatting and structure information embedded into text
  - others include LaTeX, XML, markdown
- [My favorite HTML resource](https://www.w3schools.com/html/default.asp)
- A few highlights:
  - Headers
  - Links ("anchors")
  - Tables
  - Paragraphs
- In Shiny: the UI elements product HTML, and many correspond directly to HTML elements
  - You can use `htmltools::tags` to get many HTML elements that aren't included in Shiny
    - e.g. `htmltools::tags$table`, `$tr`, `$th`, `$td` to make a table manually
    - [Documentation](https://shiny.rstudio.com/reference/shiny/1.7.4/builder.html), but the list of available tags is incomplete


## HTML only demo

<a href="bootstrap_example/index.html" target="_blank">HTML-only webpage example in `bootstrap_example/index.html`</a>

```{r html-example-code, results='asis'}
bootstrap_code <- readLines("bootstrap_example/index.html")
cat("```", sep="\n")
cat(bootstrap_code, sep="\n")
cat("```", sep="\n")
```


## CSS (1996)

- CSS = "Cascading Style Sheets"
- Language to define styles -- separate content and presentation
  - Similar to defining styles in Word, or LaTeX themes
- [My favorite CSS resource](https://www.w3schools.com/css/default.asp)
- [Example of webpage with different CSS stylesheets](https://www.w3schools.com/css/css_intro.asp)


## Where to use CSS?

- [More information here](https://www.w3schools.com/html/bootstrap_css.asp)
- Inline: in the style attribute of any HTML element
- Internal: in the style element of the HTML document `<head>` section
- External: in a standalone CSS stylesheet file
- In Shiny, you can add a `style` attribute to any UI element, e.g. `p("Some example text", style="color: blue; font-size: 20px;")`
  - This is useful if the built in styling is mostly working, but you want to make one or two small changes to it

- DEMO: open the HTML + CSS web page `css_example/index.html` in your web browser and in a text editor to see how the example changes with CSS from the `css_example/mystyle.css`


## HTML + CSS demo

<a href="css_example/index.html" target="_blank">HTML + CSS webpage example in `css_example/index.html`</a>

### CSS code

```{r css-css-example-code, results='asis'}
css_css_code <- readLines("css_example/mystyle.css")
cat("```", sep="\n")
cat(css_css_code, sep="\n")
cat("```", sep="\n")
```

### HTML code

```{r css-html-example-code, results='asis'}
css_bootstrap_code <- readLines("css_example/index.html")
cat("```", sep="\n")
cat(css_bootstrap_code, sep="\n")
cat("```", sep="\n")
```


## Boostrap (2011)

- Bootstrap is an extremely popular open source front-end framework
  - Developed by Twitter employees (long since not working there), but not sponsored by the company and in practice entirely independent
- [Official Boostrap documentation](https://getbootstrap.com)
- Bootstrap includes:
  - Responsive layout system
  - CSS style definitions, e.g. contextual colors, buttons, etc.
  - Modern-looking components, e.g. accordion, cards, modals, collapse, list groups, etc.
  - Optional JavaScript interactivity, e.g. popovers, tooltips, etc.
  - Icons
- [My favorite Bootstrap resource](https://www.w3schools.com/bootstrap/bootstrap_ver.asp)


## HTML + Bootstrap demo

<a href="bootstrap_example/index.html" target="_blank">HTML with Bootstrap webpage example in `bootstrap_example/index.html`</a>

```{r bootstrap-example-code, results='asis'}
bootstrap_code <- readLines("bootstrap_example/index.html")
cat("```", sep="\n")
cat(bootstrap_code, sep="\n")
cat("```", sep="\n")
```


## Bootstrap and Shiny

- [Mastering Shiny chapter 8](https://mastering-shiny.org/action-layout.html#bootstrap)
- Bootstrap is built into Shiny
  - e.g. modals are a direct implementation of Bootstrap 3 modals
  - e.g. `fluidRows` and `columns` use the Bootstrap 3 grid layout
- Some history and why Shiny is still using a 2-versions-old Bootstrap:
  - Shiny originally had Bootstrap 2 built into it
  - They had to migrate to Bootstrap 3 early on and it was a huge pain
  - Now the strategy is to put additional Bootstrap elements into `bslib` and use `bslib` theming to choose Bootstrap version


## `bslib`

- [`bslib` documentation](https://rstudio.github.io/bslib/index.html)
- [Towards the next gen. of Shiny UI (talk by Carson Sievert)](https://youtu.be/ZIIH8WBHlLc)
- Trang's [`bmsdash` package](https://biogit.pri.bms.com/KSR/bmsdash) implements a few BMS-styled `bslib` themes
  - Can be used in RMarkdown documents and Shiny apps


## Web browsers

- Every web browser uses its own code base to parse and process HTML, CSS, etc.
- Web browsers use their own style sheets! [Demo here](https://browserdefaultstyles.com/)
  - These set the default style if a webpage's CSS doesn't define something

### Using your browser's inspector

- Safari: 
  - One time: File > Preferences > Advanced > Check "Show Develop menu in menu bar"
  - Develop > Show Web Inspector
- Chrome: View > Developer > Developer Tools
- Edge: [link to instructions](https://learn.microsoft.com/en-us/microsoft-edge/devtools-guide-chromium/overview)


## Build a Shiny app *without* Bootstrap or any built in CSS

- [RStudio article](https://shiny.rstudio.com/articles/html-ui.html)
- Why would you want to do this?
  - Make your Shiny app look like one page integrated into a production-grade, professionally designed web site
  - You already know CSS and the Shiny defaults just keep getting in your way
  - You want to use an alternate front-end framework (not Bootstrap)
  - You like extra work??? `¯\_(ツ)_/¯`


## Takeaways

- Building a web page with HTML is like writing in plain text
  - Add HTML element tags to build your content
- Add CSS styles to customize the presentation of your HTML content
  - Add CSS rules to define which styles to apply to which HTML elements
- Bootstrap is a library of CSS + JS + icons 
  - Add Bootstrap classes to use pre-defined CSS styles and JS functions
- Beware when following online advice to add custom CSS or a custom stylesheet -- it may not work well with Bootstrap!


## Try this

- Use `bslib` to experiment with newer Bootstrap versions and additional Bootstrap elements
- Manually add a Bootstrap class to your Shiny code
- Read [Outstanding User Interfaces with Shiny](https://unleash-shiny.rinterface.com) for even more excellent information
